Global code motion of congruent instructions

### 2017-09-16 cgo/paper.tex
code formatting

\begin{verbatim}
// Compute a merge-set breadth first order.
bool constructMergeSet_1(BFSList B, JEdges JE,
                         DomLevel DL)
  repeat = false
  V = empty list of visited edges
  for n in B:
    for e in (all incoming edges to n):
      if e is in JE and e not in V:
        V(e) = true
        let tmp = Source Node of e
        let tnode = Target Node of e
        let lnode = NULL
        while (DL(tmp) >= DL(tnode)):
          MergeSet(tmp) = MergeSet(tmp)
                          U MergeSet(tnode)
                          U {tnode}
          lnode = tmp
          tmp = dom-parent(tmp)
        repeat = still_inconsistent(lnode, JE, MergeSet)
  return repeat
\end{verbatim}
\newpage
\begin{verbatim}
// Construct merge-set of each node in
// Control-Flow-Graph G of a function.
void constructMergeSet(CFG G)
  B = Breadth First Order of G
  JE = JEdges of G
  DL = List of Path length (from root) of each node in G.
  do: // Call until a fixed point is reached.
    Repeat = constructMergeSet_1(B, JE, DL)
  while (Repeat)
\end{verbatim}

@@

\begin{verbatim}
// Return true if variable A is liveout from basic-block N
bool isLiveOutUsingMergeSet(Node N, Variable A)
  if (A is defined in N)
    if (A is used outside any basic-block other than N)
      return true
    return false

  MergeSet(N) = null
  // The merge-set of N is the union of merge-sets of its successors.
  for W in successors(N)
    MergeSet(N) = MergeSet(N) U MergeSet(W)

  // Iterate over all the uses of A and see if any intersect with the merge-set of N.
  for T in users(A)
    B = basic_block(T)
    while (B != null and B != def_bb(A))
      if B in MergeSet(N) return true
      B = dom-parent(B)
  return false
\end{verbatim}

@@

\begin{verbatim}
// Compute barriers for both hoistable and sinkable
// candidates.
void compute_barriers()
  for each basic-block B in a function:
    barrier_found = false
    for each instruction I in B:
      if I does not guarantee progress:
        mark I as a barrier instruction
        barrier_found = true
        break;

    // Find the last barrier below which instructions can be sunk.
    // If there was no barrier in B, any instruction satisfying
    // other legality checks can be sunk.
    if barrier_found:
      for each instruction I in B in reverse order:
        if I does not guarantee progress:
          mark I as a sink barrier
          break
\end{verbatim}

@@

In order to be able to move an instruction that has side effects, we must
guarantee that the side-effects of that instruction before and after its
code-motion will produce the same side-effects.  For hoisting in particular, we
need to prove that on all execution paths from the hoisting point to the end of
the program the side effects appear in the same order. In general, we cannot
introduce a new computation along any path of execution without checking for
undefined behaviors. The algorithm \emph{compute\_downward\_safety} computes if
hoisting a set of instruction to a common hoisting point is downward-safe
\cite{steven1997advanced}. Since \GCM{} is only dealing with VBEs, computing
downward-safety ensures that no undefined behaviors are introduced as a result
of code-motion.
\newline
\begin{verbatim}
// Return true if a hoistable instruction is downward safe at hoist-point
bool compute_downward_safety()
  Worklist = list of all basic-blocks of hoistable instructions
  for each basic-block B in the dominator tree starting at hoist-point:
    if Worklist is empty
       return false // Path exists! not downward safe
    if B is in Worklist:
      remove B from worklist
      remove subtree with root at B // Available at B => downward safe from B
    if B is a leaf basic-block
      return false  // Path exists! not downward safe
    if B dominates hoist-point
      return false  // Back edge
  return true
\end{verbatim}

@@

\begin{verbatim}
void doGCM()
  Analyses available: Dominator Tree, DFS Numbering, Memory SSA,
  Compute DJ-graph of function
  compute mergeset based on that (using constructMergeSet function)
  compute barriers (using compute_barriers function)
  do
    Compute GVN of each expression in the function
  while (doCodeHoisting() > 0) // Repeat hoisting if any candidates were hoisted.
  doCodeSinking()
\end{verbatim}

@@

For hoisting, we find a common dominator dominating all a set of congruent
instructions and perform legality checks, as described in
Section~\ref{subsec:legality}. Often times it is not possible to hoist all the
instructions to one common dominator, due to legality constraints, e.g.,
intersecting side-effects, or profitability constraints, e.g., increasing
register pressure; in those cases, this algorithm would partition
(Section~\ref{subsec:partition}) the list of identical instructions into subsets
which can be partially hoisted to their respective common dominators.
\newpage
\begin{verbatim}
int doCodeHoisting() // Find hoistable candidates and do hoisting
  For each value number VN with 2 or more instructions (I1, I2, ..., Ik)
    sort the instructions according to DFSIn numbers
    if first two I1, I2 are hoistable and if they are downward-safe at hoist-point
          and if they are profitable to be hoisted at hoist-point
      proceed to check the ability to hoist subsequent instruction with same VN
    else
      proceed to check if I2 and I3 (if available) can be hoisted
  For each set of instructions (I1, ..., Ik) with same VN that can be hoisted:
    move the first one I1 to the hoist-point
    update the use of all other candidates (I2, ..., Ik) to refer to I1
    remove all other instructions (I2, ..., Ik)
    if I1 has memory references (a load, store, call, etc.)
      update MemorySSA of I1 and others to point to the MemorySSA reference of I1
      remove MemorySSA reference of all others which were deleted
    update statistics
  return number of hoisted instructions
\end{verbatim}

@@

\begin{verbatim}
void doCodeSinking()
  // Find sinkable instructions
  For each value number VN with 2 or more instructions
    if there are two instructions with same immediate successor as post-dominator
      if both are only used in the same PHI-node (PN) of the successor
        if both have dependencies that are available at the sink point
           // available or can be made available by rematerialization
           mark VN as sinkable

  // Sink the instructions and update statistics
  For each pair (I1, I2) of sinkable instructions
    Move I1 to the sink point
    // the sink point is just after all the PHI-nodes in the post-dominator
    update all the uses of PHI node (PN) with I1
    Remove I2 and PN
    if I1 is a memory reference:
      update MemorySSA of I1
      update MemorySSA of removed instructions to point to the MemorySSA reference of I1
      remove MemorySSA reference of all others which were deleted
    update statistics

>>>>

\begin{verbatim}
// Compute a merge-set breadth first order.
bool constructMergeSet_1(BFSList B, JEdges JE,
                         DomLevel DL)
  repeat = false
  V = empty list of visited edges
  for n in B:
    for e in (all incoming edges to n):
      if e is in JE and e not in V:
        V(e) = true
        let tmp = Source Node of e
        let tnode = Target Node of e
        let lnode = NULL
        while (DL(tmp) >= DL(tnode)):
          MergeSet(tmp) = MergeSet(tmp)
                          U MergeSet(tnode)
                          U {tnode}
          lnode = tmp
          tmp = dom-parent(tmp)
        repeat = still_inconsistent(lnode, JE, MergeSet)
  return repeat
\end{verbatim}
\newpage
\begin{verbatim}
// Construct merge-set of each node in
// Control-Flow-Graph G of a function.
void constructMergeSet(CFG G)
  B = Breadth First Order of G
  JE = JEdges of G
  DL = Path lengths (from root) of each node in G
  do: // Call until a fixed point is reached.
    Repeat = constructMergeSet_1(B, JE, DL)
  while (Repeat)
\end{verbatim}

@@

\begin{verbatim}
// Return true if variable A is liveout from N
bool isLiveOutUsingMergeSet(Node N, Variable A)
  if (A is defined in N)
    if (A is used outside any basic-block other than N)
      return true
    return false

  MergeSet(N) = null
  // The merge-set of N is the union of
  // merge-sets of its successors.
  for W in successors(N)
    MergeSet(N) = MergeSet(N) U MergeSet(W)

  // Iterate over all the uses of A and see
  // if any intersect with the merge-set of N.
  for T in users(A)
    B = basic_block(T)
    while (B != null and B != def_bb(A))
      if B in MergeSet(N) return true
      B = dom-parent(B)
  return false
\end{verbatim}

@@

\begin{verbatim}
// Compute barriers for both hoistable and sinkable
// candidates.
void computeBarriers()
  for each basic-block B in a function:
    barrier_found = false
    for each instruction I in B:
      if I does not guarantee progress:
        mark I as a barrier instruction
        barrier_found = true
        break;

    // Find the last barrier below which instructions
    // can be sunk. If there was no barrier in B, any
    // instruction satisfying other legality checks
    // can be sunk.
    if barrier_found:
      for each instruction I in B in reverse order:
        if I does not guarantee progress:
          mark I as a sink barrier
          break
\end{verbatim}

@@

In order to be able to move an instruction that has side effects, we must
guarantee that the side-effects of that instruction before and after its
code-motion will produce the same side-effects.  For hoisting in particular, we
need to prove that on all execution paths from the hoisting point to the end of
the program the side effects appear in the same order. In general, we cannot
introduce a new computation along any path of execution without checking for
undefined behaviors. The algorithm \emph{compute\_downward\_safety} computes if
hoisting a set of instruction to a common hoisting point is downward-safe
\cite{steven1997advanced}. Since \GCM{} is only dealing with VBEs, computing
downward-safety ensures that no undefined behaviors are introduced as a result
of code-motion.
\newline
\begin{verbatim}
// Return true if a hoistable instruction is downward
// safe at hoist-point
bool compute_downward_safety(Node HP)
  Worklist = basic-blocks of hoistable instructions
  DT = The dominator tree
  HP = the hoist-point
  for each basic-block B in the DT starting at HP:
    if Worklist is empty
       return false // Path exists! not downward safe
    if B is in Worklist:
      remove B from worklist
      // Available at B => downward safe from B
      remove subtree with root at B
    if B is a leaf basic-block
      // Path exists! not downward safe
      return false
    if B dominates hoist-point
      return false  // Back edge
  return true
\end{verbatim}

@@

\begin{verbatim}
void doGCM()
  Analyses available:
    Dominator Tree, DFS Numbering, Memory SSA
  Compute DJ-graph of function
  constructMergeSet(CGF)
  computeBarriers()
  do:
    Compute GVN of each expression in the function
   // Repeat hoisting if any candidates were hoisted.
  while (doCodeHoisting() > 0)
  doCodeSinking()
\end{verbatim}

@@

For hoisting, we find a common dominator dominating all a set of congruent
instructions and perform legality checks, as described in
Section~\ref{subsec:legality}. Often times it is not possible to hoist all the
instructions to one common dominator, due to legality constraints, e.g.,
intersecting side-effects, or profitability constraints, e.g., increasing
register pressure; in those cases, this algorithm would partition
(Section~\ref{subsec:partition}) the list of identical instructions into subsets
which can be partially hoisted to their respective common dominators.
\newpage
\begin{verbatim}
// Find hoistable candidates and do hoisting
int doCodeHoisting()
  For each value number VN with 2 or more instructions
    sort the instructions according to DFSIn numbers
    if first two I1, I2 are hoistable
        && downward-safe at hoist-point
        && profitable to be hoisted at hoist-point:
      continue the check for subsequent
        instructions with same VN
    else:
      proceed to check if I2 and I3 (if available)
      can be hoisted

  For each set of hoistable instructions:
    move the first one (I1) to the hoist-point
    update their users others to refer to I1
    remove all other instructions (I2, ..., Ik)
    if I1 has memory references:
      update MemorySSA of I1 and others
    update statistics
  return number of hoisted instructions
\end{verbatim}

@@

\begin{verbatim}
int doCodeSinking()
  // Find sinkable instructions
  For each value number VN with 2 or more instructions:
    if there are two instructions with same immediate
        successor(S) as post-dominator:
      if both are only used in the same PHI-node of S
          && dependencies of both are available at S:
         mark instructions as sinkable

  // Sink the instructions and update statistics
  For each pair (I1, I2) of sinkable instructions:
    Move I1 to the sink point
    // the sink point is just after all
    // the PHI-nodes in the post-dominator
    update all the uses of PHI node (PN) with I1
    Remove I2 and PN
    if I1 is a memory reference:
      update MemorySSA of I1
      update MemorySSA of removed instructions
        to point to the MemorySSA reference of I1
      remove MemorySSA reference of all
        others which were deleted
    update statistics
